\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-13 article.tex
more detail

Again the \texttt{OCamlgraph} library is used, but this time \texttt{Imperative.Graph.Abstract} is parameterized over the aforementioned set type \texttt{S.SS.t}. % FIXME: whyyy????
Such that several variables can be assigned to the same node (more on this in section \ref{coalescing}). A helper function is defined, that looks up a variable's symbol in the symbol table over vertices and adds it if it doesn't exist. Otherwise it simply returns the existing vertex identifier:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let vert g e t =
  match S.ST.find_opt e t with
  | Some v -> (v, t)
  | None ->
      let v = S.SS.add e S.SS.empty |> G.V.create in
      G.add_vertex g v;
      (v, S.ST.add e v t)
\end{minted}
Then the interference graph is derived in Figure \ref{fig:interf}. First, all parameters are connected, then the instructions are processed according to the book.

\begin{figure}[H]
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let interf (params : Ll.uid list) (insns : Cfg.insn list) _ (out : S.SS.t array) =
  let g = G.create () in
  let vert2 t e = vert g e t |> snd and vert3 e t = vert g e t |> snd in
  let edge s1 s2 t =
    let v1, t = vert g s1 t in
    let v2, t = vert g s2 t in
    if v1 <> v2 then G.add_edge g v1 v2;
    t
  in
  let t = List.fold_left vert2 S.ST.empty params in
  let param t p1 =
    let param t p2 = if p1 <> p2 then edge p1 p2 t else t in
    List.fold_left param t params
  in
  let t = List.fold_left param t params in
  let setverts t s = S.SS.fold vert3 s t in
  let t = List.map (def S.SS.empty) insns |> List.fold_left setverts t in
  let defoutedges t (i, n) =
    let defs = def S.SS.empty n in
    let outedge e1 t = S.SS.fold (edge e1) out.(i) t in
    S.SS.fold outedge defs t
  in
  let t = List.mapi (fun i n -> (i, n)) insns |> List.fold_left defoutedges t in
  (t, g)
\end{minted}
\caption{Overall implementation of \texttt{interf}\label{fig:interf}}
\end{figure}

@@

The \mintinline{llvm}{call} operation is by far the most complicated one. Mostly because of calling conventions,
The \texttt{\%rax} register is zeroed before every call as its lowermost byte register \texttt{\%al} is used to indicate the number of variadic arguments \cite[25]{sysv}.

@@

The testing framework is roughly based on the build system that came with the Tiger development environment. It is based on a collection of files that can be compiled and executed on their own. Each of these are enumerated in the test runner (\texttt{bin/testrunner.ml}) with parameters associated.   Each of these are then compiled with the \texttt{clang} compiler and executed, capturing all the relevant output, which is stdout and exit code. After this, the same test file is compiled for each of the allocators. The executables generated are then also executed, also storing the output data.   All test cases are deterministic and will be  halted by configurable timeout at some point. This is necessary because tests can easily end in an infinite loop because of erroneous translation. As each of these halt or time out, their output is compared to what was output from executables generated by the 'correct' compiler of \texttt{clang}.

@@

Each \lstinline!.ll! file present in the \lstinline!benches! directory is listed here in a table denoting the input paramater on the left column and type of allocator used.

>>>>

Again the \texttt{OCamlgraph} library is used, but this time \texttt{Imperative.Graph.Abstract} is parameterized over the aforementioned set type \texttt{S.SS.t}. % FIXME: whyyy????
Such that several variables can be assigned to the same node (more on this in section \ref{coalescing}). A helper function is defined, that looks up a variable's symbol in the symbol table over vertices and adds it if it doesn't exist. Otherwise it simply returns the existing vertex identifier:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let vert g e t =
  match S.ST.find_opt e t with
  | Some v -> (v, t)
  | None ->
      let v = S.SS.add e S.SS.empty |> G.V.create in
      G.add_vertex g v;
      (v, S.ST.add e v t)
\end{minted}
Then the interference graph is derived in Figure \ref{fig:interf}. First, all parameters are connected, then the instructions are processed as described in the Appel text \cite[222]{tiger}: phi nodes, as they are the only \texttt{mov} instructions, interference edges are added with all  \(\mathit{out}[n] \setminus \{b_1, ..., b_n\}\) (see \eqref{phi}) while all others are just \(\mathit{out}[n]\).

\begin{figure}[H]
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let interf (params : Ll.uid list) (insns : Cfg.insn list) _ (out : S.SS.t array) =
  let g = G.create () in
  let vert2 t e = vert g e t |> snd and vert3 e t = vert g e t |> snd in
  let edge s1 s2 t =
    let v1, t = vert g s1 t in
    let v2, t = vert g s2 t in
    if v1 <> v2 then G.add_edge g v1 v2;
    t
  in
  let t = List.fold_left vert2 S.ST.empty params in
  let param t p1 =
    let param t p2 = if p1 <> p2 then edge p1 p2 t else t in
    List.fold_left param t params
  in
  let t = List.fold_left param t params in
  let setverts t s = S.SS.fold vert3 s t in
  let t = List.map (def S.SS.empty) insns |> List.fold_left setverts t in
  let ids = function Ll.Id id -> Some id | _ -> None in
  let defoutedges t (i, n) =
    let defs = def S.SS.empty n in
    let out =
      match n with
      | Cfg.Insn (_, Ll.PhiNode (_, ops)) ->
          List.map fst ops |> List.filter_map ids |> S.SS.of_list
          |> S.SS.diff out.(i)
      | _ -> out.(i)
    in
    let outedge e1 t = S.SS.fold (edge e1) out t in
    S.SS.fold outedge defs t
  in
  let t = List.mapi (fun i n -> (i, n)) insns |> List.fold_left defoutedges t in
  (t, g)
\end{minted}
\caption{Overall implementation of \texttt{interf}\label{fig:interf}}
\end{figure}

@@

The \mintinline{llvm}{call} operation is by far the most complicated one. Mostly because of calling conventions, caller saved registers listed in Table \ref{tab:trans} need to be saved/restored before/after all. This isn't necessary in all cases, only the cases where they are actually used (which ought to be a simple lookup in the assignments table), but is not accounted for in this implementation. Instead they are all saved/restored regardless of use this could also be further optimized by simply subtracting from \texttt{\%rsp} and moving them to their respective stack locations manually,  instead of using push/pop.
Additionally, the \texttt{\%rax} register is zeroed before every call as its lowermost byte register \texttt{\%al} is used to indicate the number of variadic arguments \cite[25]{sysv}. This isn't actually accounted for, instead it is just set to zero regardless.
Furthermore, the way in which arguments are passed is also unnecessarily convoluted, but stems from a corner case in which variables were overwritten by moving to the respective register of the \(n\) function argument. This can be remedied by ordering the \texttt{mov}s correctly, and even in the case where there are cycles and \texttt{mov}s cannot be made without overwriting, one of the scratch registers can be used instead. But as a workaround for this, the arguments are simply pushed from the source operands and popped to their respective argument registers (in reverse order).

@@

The testing framework is roughly based on the build system that came with the Tiger development environment. It is based on a collection of files that can be compiled and executed on their own. Each of these are enumerated in the test runner (\texttt{bin/testrunner.ml}) with parameters associated.   Each of these are then compiled with the \texttt{clang} compiler and executed, capturing all the relevant output, which is stdout and exit code. After this, the same test file is compiled for each of the allocators. The executables generated are then also executed, also storing the output data.   All test cases are deterministic and will be  halted by configurable timeout at some point. This is necessary because tests can easily end in an infinite loop because of erroneous translation. As each of these halt or time out, their output is compared to what was output from executables generated by the 'correct' compiler of \texttt{clang}.
Though testing the outputs of several stages much like the Tiger compiler from the compilers course, this wasn't realized in time.
All of the test entries in \texttt{bin/testrunner.ml} were either working at one point and the result of a regression or it was a work in progress that wasn't realized in time either. Except the \~ 20\% of the LLVM test suite that were never working.

@@

Some of the \texttt{.ll} file present in the \texttt{benches} directory are listed here in a table denoting the input paramater on the left column and type of allocator used.
The data measured is extracted from the \texttt{perf} as mentioned above. This is done in \texttt{bin/bench.ml} which also contains the commands and parameters to execute each benchmark.
